Building Suitable Datasets for Soft Computing and Machine Learning Techniques from Meteorological Data Integration: A Case Study for Predicting Significant Wave Height and Energy Flux

### 2021-01-11 latex_manuscript/MDPI/2_energies_special_issue/envio-3/coverLetter-responseToReviewers/cover-letter-response-reviewers-energies-1069865.tex

\newcommand{\myTitle}{Building Suitable Datasets for Soft Computing and Machine Learning Techniques from Meteorological Data Integration: A Case Study for Predicting Significant Wave Height and Energy Flux}

\newcommand{\myShortTitle}{Response to reviewers}

\newcommand{\myJournal}{\textbf{Energies}}

\newcommand{\myRefPaper}{\textbf{Paper reference: energies-1069865}}

@@

\thispagestyle{plain}

\begin{center}
	{\LARGE\myTitle} \vspace{0.5cm} \\
	{\LARGE\myJournal} \vspace{0.5cm} \\
	{\Large\myRefPaper} \vspace{0.5cm} \\
% 	\today \vspace{0.5cm} \\
	\myAuthors \vspace{0.5cm} \\
	{\myDepartament} \vspace{0.5cm} \\
	{\myEmail}
\end{center}

\tableofcontents

\section{Responses to the Editor}
We would like to thank the Editor and the anonymous reviewers for this new round of review. We have addressed the final points required by the reviewers.

@@

%\textcolor{red}{Summary of changes}

Please find below the responses to the reviewers' comments.

\section{Responses to reviewers' comments}
We sincerely thank the anonymous reviewers for their careful reviews, constructive comments and hints to improve the impact of this work.

Most important changes highlighted in \textcolor{blue}{blue color} in the previous version of the manuscript have already been set to the default color (\textbf{black color}). Lines highlighted in \textcolor{blue}{blue color} correspond to the final points of this new revision. Please, find below detailed responses to the comments.

% Additionally, in order to improve the English quality and style we have carried out a proof-reading.
% Additionally, in order to improve the paper's readability, we have carried out a proof-reading, where different typos and weird expressions have been corrected.
%Additionally, in order to improve the english language and style we have carried out a review.

@@

% The most important changes made in the first revision of the paper concerning the initial version are  \textcolor{blue}{marked in blue}, while the changes made in this second revision  \textcolor{red}{are marked in red}. Also, this letter response attempts to clarify the questions asked by the reviewers.

%\newpage
\section{Reviewer \#1}
\addtocounter{section}{-2}

%%%%%%%%%%%%%%%%%%%%%%%%%% Comment %%%%%%%%%%%%%%%%%%%%%%%%%%
\rcomment
{
Dear authors,
Thank you for your efforts in revising the manuscript. I believe you did a great job in the revision and the changes alleviated my concerns regarding the manuscript. Therefore, I recommend its publication.

\vspace{0.5cm}
Looking forward to seeing the published version of your paper.

\vspace{0.5cm}
Good luck
}

\textbf{Response:}
{
Thank you very much for the excellent review carried out to our work.
}
%%%%%%%%%%%%%%%%%%%%%%%%%% Comment %%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%% Another Reviewer %%%%%%%%%%%%%%%%%%%%
%\newpage
\clearpage
\addtocounter{section}{+2}
\section{Reviewer \#2}
\addtocounter{section}{-2}

%%%%%%%%%%%%%%%%%%%%%%%%%% Comment %%%%%%%%%%%%%%%%%%%%%%%%%%
\rcomment{
The manuscript is in a good shape. The manuscript is suitable for publication in its current form.
}

\textbf{Response:}
{
We would like to thank you again for the excellent review carried out to our work.
}
%%%%%%%%%%%%%%%%%%%%%%%%%% Comment %%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%% Another Reviewer %%%%%%%%%%%%%%%%%%%%
%\newpage
\clearpage

\addtocounter{section}{+2}
\section{Reviewer \#3}
\addtocounter{section}{-2}

%%%%%%%%%%%%%%%%%%%%%%%%%% Comment %%%%%%%%%%%%%%%%%%%%%%%%%%
\rcomment{
The manuscript is related to an interesting topic and based on open database.

\vspace{0.5cm}
The structure is organized in a good way. Few adjustments are requested to introduction and conclusions.

\vspace{0.5cm}
First of all, in the abstract openess of data is mentioned but it is not discussed deeply in stating the background. See recent articles published in top-ranked Journals such as Energy, Energies and others. As example, consider

\vspace{0.5cm}
https://doi.org/10.1016/j.energy.2020.118803

\vspace{0.5cm}
https://doi.org/10.3390/en13226095

\vspace{0.5cm}
The research question formulated by the authors to fill the identified research gap should open the conclusions section as
well as reporting limitation and replicability potential of the answer.

\vspace{0.5cm}
Please check carefully English.
}

@@

\textbf{Response:}
{
First of all, we would like to thank Reviewer \#3 for his/her suggestions.

We have mentioned the two sources of data from which the data integration using SPAMDA is carried out in \textbf{Section ''1. Introduction`` (lines 117-122 - ''The meteorological data ... build datasets``)}, but we have not made an extensive explanation of these data because they are well known. Data provided by the NDBC and the NNRP are available for the scientific community and there are manuals and detailed descriptions on the websites of both information sources. Even so, we have dedicated \textbf{Section ``2. Meteorological data sources''} to briefly discuss both sources of data and how they are used in SPAMDA, supporting this section with bibliographical references and website addresses. 
%Figure 1 and Table 1 are also included as additional explanations of both data sources.

We sincerely believe that the paper [11] indicated by the reviewer underlines the importance of having well-formatted and high-quality data to be able to use them with modelling techniques, as well as the importance of both data and modelling tools being available to the scientific community and decision-makers. This reinforces our work, so we have introduced this reference in \textbf{Section ''1. Introduction`` (lines 37-40 - ''In this sense ... decision-makers
[11]``)}. The reference [25] also provides background to our work, since it reflects the importance of the correct arrangement of WECs for a better obtaining of power output and energy. This reference has been introduced in \textbf{Section ''1. Introduction`` (line 69)}.

[11] Manfren, M.; Nastasi, B.; Groppi, D.; Astiaso Garcia, D. Open data and energy analytics - An analysis of essential information for energy system planning, design and operation. Energies 2020, 13(9), 2334.

[25] Amini, E.; Golbaz, D.; Amini, F.; Majidi Nezhad, M.; Neshat, M.; Astiaso Garcia, D. A Parametric Study of Wave Energy Converter Layouts in Real Wave Models. Energies 2020, 13(22), 6095.

Finally, we have modified Section ''5. Conclusions`` to begin with the research gap. The replicability of data used in Section ''4. A case study applied to Gulf of Alaska`` or any other dataset created with SPAMDA, can be obtained by providing the parameters entered in the tool or simply by providing the final dataset after the last data integration step. These clarifications have been included in \textbf{Section ''5. Conclusions`` (lines 755-763 - ''Studies on marine energy ... among others``, lines 780-783 - ''The final datasets ... SC or ML tool used``)}.

%As a conclusion, we believe that open data, open science, open innovation and, more specifically, open energy modelling principles are crucial to create an effective science-policy-market interaction in energy and sustainability transitions. Research efforts in energy modelling should be oriented to the identification of solutions that represent good compromises between connectivity and diversity and to the definition of appropriate hierarchies of data and models, for example by using linked open data schema and meta-models.

### 2021-01-11 latex_manuscript/MDPI/2_energies_special_issue/envio-3/latex/energies-1069865.tex
Version final para el tercer envio Special Issue Energies

%=================================================================
% Full title of the paper (Capitalized)
\Title{Building Suitable Datasets for Soft Computing and Machine Learning Techniques from Meteorological Data Integration: A Case Study for Predicting Significant Wave Height \textcolor{blue}{and Energy Flux}}

@@

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{Meteorological data are extensively used to perform environmental learning. Soft Computing (SC) and Machine Learning (ML) techniques represent a valuable support in many research areas, but require datasets containing information related to the topic under study. Such datasets are not always available in an appropriate format and its preparation and pre-processing implies a lot of time and effort by researchers. This paper presents a novel software tool with an user-friendly GUI to create datasets by means of management and data integration of meteorological observations from two well-known sources of information: the \textit{National Data Buoy Center} and the \textit{National Centers for Environmental Prediction and for Atmospheric Research Reanalysis Project}. Such datasets can be created using buoys and reanalysis data through customisable procedures, in terms of temporal resolution, predictive and objective variables, \textcolor{blue}{ and can be used by SC and ML methodologies for prediction tasks (classification or regression). The objective is providing the research community with an automated and versatile system for the casuistry that entails well-formed and quality data integration, potentially leading to better prediction models. The software tool can be used as a supporting tool for coastal and ocean engineering applications, sustainable energy production or environmental modelling; as well as for decision making in the design and construction of coastal protection structures, marine transportation, offshore industry, ocean energy converters and efficient operation of offshore and coastal engineering activities. Finally, to illustrate the applicability of the proposed tool, a case study to classify waves depending on their significant height and to predict energy flux in the Gulf of Alaska is presented. }
%Therefore, it does not make a validation or comparison study of the performance of state-of-the-art prediction algorithms.
}

@@

		A better understanding of the environment is of vital importance for science, contributing not only to more efficient exploitation of natural resources but also to the development of new strategies aimed at its protection. In that sense, meteorological observations provide an essential and valuable source of information which is widely used by researchers to address environmental learning, comprehension, prediction and conservation in numerous oceanic and atmospheric studies of a wide variety of areas (e.g. energy, climate change, agriculture, etc.). Some specific examples of the diversity of fields in which meteorological data can be used in are, among others: estimation of global solar radiation based on sunshine duration \cite{SHAHRUKHANIS2019179}, directional analysis of sea storms \cite{LAFACE201545}, estimation of hybrid energy systems taking into account economic and environmental objectives\cite{Kumar2020}, wind power ramp events prediction \cite{DORADOMORENO2017428}, sea surface temperature prediction \cite{He2020}, study of the responses exhibited by plankton to fluid motions \cite{FUCHS2016109}, trends in solar radiation \cite{SILVA20101852} or simulation of extreme near shore sea conditions \cite{GOULDBY201415}. All these studies require a prior data collection and its adaptation to a specific format that allows the interpretation of them.

		Once quality and well-formed data are obtained, these can be used to extract information and build prediction models that explain the behavior of a certain problem. \textcolor{blue}{The choice of the appropriate model, either in engineering problems or in any other problem, is also an important factor in addition to data \cite{Alizadeh2019, Alizadeh2020}}. Although Soft Computing (SC) and Machine Learning (ML) techniques have the ability to handle uncertainty in data and are extensively used for modelling purposes, the real challenge in modelling studies is due to the inadequacy of data, since the adequacy of the models depend mainly on the the quality of the information used, so that if a researcher does not have quality data there will be no quality models.

		Continuing with this line, special purpose software is usually developed to help researchers to advance in their studies related to energy and environmental modelling, becoming a great support for decision-making in the exploitation and protection of the environment. In \cite{Neeraj2020}, a software package in R called ``ForecastTB'' is developed to compare the accuracy of different prediction methods as related to the characteristics of a time series dataset, presenting the software as a stepping stone in ML automation modelling.
		In \cite{LO2015293}, an integrated simulation tool for the optimum design of bifacial solar panel with reflectors is presented. This tool can also be used to analyse the performance of the solar cells. A framework for data integration of offshore wind farms is implemented in \cite{NGUYEN2013150} in order to facilitate data exchange and improve operation and maintenance practices. 
		In \cite{Roberta2020}, a new software tool named ``Storage LCA Tool`` is presented for comparing PCM - phase change materials- storage systems  with traditional systems that do not entail energy storage, being beneficial in order to support decision-making on energy concepts for buildings.
		A risk assessment tool to improve safety standards and emergency management in onshore wind farms, is presented in \cite{ASTIASOGARCIA201648}. Raabe et al. \cite{RAABE2010213} developed two software tools, \textit{Model of Equilibrium of Bay Beaches} (MEPBAY) and \textit{Coastal Modelling System} (SMC), to support different operational levels of headland-bay beach in coastal engineering projects, and Motahhir et al. \cite{MOTAHHIR20199} developed an open hardware/software test bench for solar tracker.

		Marine energy prediction is currently a hot topic where meteorological data is used in. Marine Renewable Energy (MRE) is one of the most important renewable and sustainable energy sources available in our environment \cite{en12050787}, and it includes ocean thermal energy, marine tidal current energy and wave energy, among others. Its benefits and great potential \cite{ZEYRINGER20181281} make it one of the most relevant natural resources, playing a crucial role not only in the reduction of the emission of greenhouse gases but also in all other aspects involved in the difficult challenge of the transition to a low carbon footprint society \textcolor{blue}{\cite{en12091657, BREDE201344, Alizadeh2020a}}. Wave energy exhibits a more stable power supply than wind energy and even solar energy. In recent years \textit{Wave Energy Converters} (WECs) \cite{FALCAO2010899} have been developed and widely installed to transform this wave energy into electricity, which can be injected into the electric network or supplied to existing offshore oil and gas platforms \cite{OLIVEIRAPINTO2019556} or seawater desalination plants \cite{FERNANDEZPRIETO2019546}, among others. WECs are mechanical devices that convert kinetic energy into electrical energy by means of either the vertical oscillation of waves or the linear motion of them. Nevertheless, waves are difficult to be characterised due to their stochastic nature, because of the influence of a large number of environmental factors that exert on them \cite{ochi1998}. As a consequence of this complexity, many aspects of WEC design, deployment and operation \textcolor{blue}{\cite{CROWLEY2018159, Abdelkhalik2016, 6898109}} need a proper prediction of waves \textcolor{blue}{\cite{en11010011, Kaloop2020}}, in order to maximise the wave energy extraction \cite{en80910370}. For this purpose WECs use wave \textit{flux of energy} ($F_e$) which can be calculated from the two most important wave parameters in this regard: \textit{significant wave height} ($H_s$) and \textit{wave energy period} ($T_e$).

		Currently, and as a support to traditional study procedures, SC and ML techniques \cite{Sang-Yong2014,Bishop:2006:PRM:1162264} are being widely used in numerous research fields related to classification, regression and optimisation tasks, obtaining significant improvements in the performance of the results, \textcolor{blue}{ either in engineering \cite{Alizadeh2020}, energy or environmental problems} \cite{Fi-John2019,Mosavi2019,GUO201816}. SC and ML methodologies can be used not only by experienced computer scientists but also by other researchers. For example, the well-known \textit{Waikato Environment for Knowledge Analysis} (WEKA) \cite{WEKA} software tool provides researchers with a wide collection of ML  algorithms. ML techniques have been already applied to tackle wave characterisation, accurately estimating $H_{s}$ and $T_{e}$ parameters \cite{DURANROSAL2017268, KUMAR2017605}, given that robustness of ML methods can tackle the previously explained difficulties in wave energy prediction. \textcolor{blue}{In \cite{Ali2020}, a reliable ML model based on multiple linear regression and covariant-weighted least square estimation for $H_s$ modelling is presented in order to forecast half-hourly nearly real-time significant wave height values. In \cite{Cornejo-Bueno2016}, an  approach for feature selection problems is developed and applied to a specific problem of $H_s$ and $F_e$ prediction in oceanic buoys,  obtaining excellent results in terms of prediction quality. In \cite{Emmanouil2020}, a Bayesian Network system provides an useful tool for the decision making process of installation and maintenance operations in offshore wind farms using predictions of $H_s$, among others. In \cite{Shamshirband2020}, several ML methods are implemented and compared for the prediction of $H_s$ in the Persian Gulf, the extreme learning machine (ELM) providing the best results.} The problem is that, in order to apply \textcolor{blue}{ML and SC techniques}, it is essential to obtain datasets with relevant information about the issue under study, used to infer knowledge. Usually, these datasets are not publicly available in a friendly format, and their generation is the first step needed.

		The information to create these datasets related with MRE can be obtained from meteorological observations, but such information may be available in a inappropriate format and even contain missing values or measurements. Consequently, it is usually required to perform pre-processing tasks for improving the quality of the data, such as the replacement of missing values, outlier detection or data normalisation, among others. Furthermore, if more than one source of information is used to achieve a better characterisation of the problem under study \cite{JOHANSSON2015143, FERNANDEZ201544, Adams2010}, then a data integration process, denominated as the matching process in this document, has to be carried out by researchers to manually create the datasets with the needed information. Given that such process is of great relevance and has an extensive casuistry, the present work has been specially focused on it. Moreover, depending on the subject and the SC and ML technique to be applied, or even if the researcher considers other factors in order to improve the results or have more in-depth conclusions, the datasets would have to be updated afterwards. In summary, many important details and different intermediate steps have to be considered when creating suitable datasets, specially when data integration is required, resulting in an extremely tedious task.

@@

			\item It makes the researcher focus on environmental modelling, without having to worry about the development of scripts or mechanical tasks, avoiding laborious pre-processing procedures, that imply a lot of time and effort in early stages of the research.

@@

			\item It provides information about the quality and quantity of the data. \textcolor{blue}{SPAMDA allows preliminary studies of missing values (dates or measurements not recorded) in buoys managed by NDBC, so that the researcher can have an idea of the quality of the data recorded by the buoys and about their suitability for the intended purpose. In any case, SPAMDA allows data integration taking into account such missing values when needed by the user.}

			\item \textcolor{blue}{Estimation of the amount of energy flux that can be produced at different prediction horizons: short-term, mid-term or long-term. Although this work does not focus on models performance, it should be taken into account that models tend to generalise worse with greater prediction horizons.}

@@

			\item Possibility of selecting one or more reanalysis nodes near the localisation under study, \textcolor{blue}{which could provide a better description of the problem to achieve more accurate models.}

			\item \textcolor{blue}{Although pre-processing is not the main objective of SPAMDA, the tool also provides some basic pre-processing filters on buoys measurements, such as normalisation and missing data recovery.}

			\item It facilitates data management and well-organised storage of the datasets. \textcolor{blue}{Environmental studies in different geographical locations can be carried out by merely introducing and using other collected data.}

@@

		\textcolor{blue}{Therefore, the functionalities and characteristics that SPAMDA offers make it a supporting tool for researchers, which could be used in applications related to coastal and ocean engineering, and also in marine energy prediction. In \cite{Kumar2020}, the estimation of energy supply sources in hybrid energy systems is based on the amount of energy that can be obtained by a marine energy system within a prediction horizon. Regulation of WECs to avoid malfunction or breakage, depending on the significant wave height and/or energy flux expected, as well as the possibility of reconfiguring them in order to maximise the wave energy extraction, is studied in \cite{CROWLEY2018159, Abdelkhalik2016}. The prediction of the energy that could be obtained from a certain maritime location is considered in \cite{OLIVEIRAPINTO2019556, FERNANDEZPRIETO2019546} in order to know whether it is convenient to install WECs as power supply in marine structures, such as offshore oil and gas platforms or seawater desalination plants. In \cite{Ali2019}, significant wave height forecasting is applied for decision-making in exploitation and environmental protection for the construction of marine energy storage plants, future strategies on renewable energy and coastal planning. Other examples of application are: design of offshore structures and ports \cite{CHATZIIOANNOU2017126}, decision-making and risk assessment about operational works in the sea \cite{DALGIC2015211}, security systems for structures or naval security \cite{Spaulding2020}.}

@@

			\item NDBC is a part of the \textit{National Weather Service} (NWS). NDBC designs, develops, operates, and maintains a network of data collecting buoys (stations). The mission of the network is to collect real-time marine meteorological and oceanographic observations, such as $H_s$, dominant wave period, or wind speed and direction, among others.

			The buoys maintained by NDBC are deployed in the coastal and offshore waters around oceans and seas, and they are equipped with assorted sensors which allow them to perform different measurements. The information collected by the buoys is available on the NDBC website \cite{NOAA_1}, and it is divided into different groups. One of them corresponds to standard meteorological information of the historical data collected by each buoy, which can be downloaded as annual text files and whose format was adopted by NDBC since January 2007 \cite {NOAA_2}. These files contain hourly measurements per day from $00$:$50$ to $23$:$50$ UTC (Universal Time Coordinated) and from $23$:$50$ 31th December of the previous desired year to $22$:$50$ 31th December of the desired year. In Table \ref{tab:measurementsDescription}, a comprehensive measurement description and the corresponding units are provided as a summary for the reader. A fragment of one of these files, which contains the measurements collected during year $2017$ by the buoy identified as \textit{Station 46001} in NDBC, is shown in Fig. \ref{fig:fragmentAnnualTexFile}. Each column corresponds to a meteorological variable or attribute, and each row or instance corresponds to the values of the measurements collected by the buoy for each attribute at a specific date and time.

@@

				\caption{Measurements descriptions and units of each meteorological variable or attribute collected by the buoys.}
				\label{tab:measurementsDescription}
				\footnotesize
				\centering

				\begin{tabular}{ccm{8.60cm}@{\setlength{\tabcolsep}{0pt}}m{0.0cm}}

@@

					WDIR & degT & Wind direction (the direction the wind is coming from in degrees clockwise from true North) during the same period used for WSPD. \\

					\cellcolor{gray090}WSPD & \cellcolor{gray090} m/s & \cellcolor{gray090} Wind speed (m/s) averaged over an eight-minute period for buoys and a two-minute period for land stations. Reported Hourly. \\

					GST &  m/s & Peak 5 or 8 second gust speed (m/s) measured during the eight-minute or two-minute period. \\

					\cellcolor{gray090} WVHT & \cellcolor{gray090} m & \cellcolor{gray090} Significant wave height (meters) is calculated as the average of the highest one-third of all of the wave heights during the 20-minute sampling period. \\

					DPD & sec & Dominant wave period (seconds) is the period with the maximum wave energy. \\

					\cellcolor{gray090}APD & \cellcolor{gray090} sec & \cellcolor{gray090} Average wave period (seconds) of all waves during the 20-minute period. \\

					MWD & degT & The direction from which the waves at the dominant period (DPD) are coming. The units are degrees from true North, increasing clockwise, with North as 0 (zero) degrees and East as 90 degrees. \\

>>>>

%=================================================================
% Full title of the paper (Capitalized)
\Title{Building Suitable Datasets for Soft Computing and Machine Learning Techniques from Meteorological Data Integration: A Case Study for Predicting Significant Wave Height and Energy Flux}

@@

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{Meteorological data are extensively used to perform environmental learning. Soft Computing (SC) and Machine Learning (ML) techniques represent a valuable support in many research areas, but require datasets containing information related to the topic under study. Such datasets are not always available in an appropriate format and its preparation and pre-processing implies a lot of time and effort by researchers. This paper presents a novel software tool with an user-friendly GUI to create datasets by means of management and data integration of meteorological observations from two data sources: the \textit{National Data Buoy Center} and the \textit{National Centers for Environmental Prediction and for Atmospheric Research Reanalysis Project}. Such datasets can be created using buoys and reanalysis data through customisable procedures, in terms of temporal resolution, predictive and objective variables, and can be used by SC and ML methodologies for prediction tasks (classification or regression). The objective is providing the research community with an automated and versatile system for the casuistry that entails well-formed and quality data integration, potentially leading to better prediction models. The software tool can be used as a supporting tool for coastal and ocean engineering applications, sustainable energy production or environmental modelling; as well as for decision making in the design and building of coastal protection structures, marine transport, ocean energy converters and well-planned running of offshore and coastal engineering activities. Finally, to illustrate the applicability of the proposed tool, a case study to classify waves depending on their significant height and to predict energy flux in the Gulf of Alaska is presented.}

@@

		A better understanding of the environment is of vital importance for science, contributing not only to more efficient exploitation of natural resources but also to the development of new strategies aimed at its protection. In that sense, meteorological observations provide an essential and valuable source of information which is widely used by researchers to address environmental learning, comprehension, prediction and conservation in numerous oceanic and atmospheric studies of a wide variety of areas (e.g. energy, climate change, agriculture, etc.). Some specific examples of the diversity of fields in which meteorological data can be used in are, among others: global solar radiation estimation \cite{SHAHRUKHANIS2019179}, directional analysis of sea storms \cite{LAFACE201545}, estimation of hybrid energy systems taking into account economic and environmental objectives\cite{Kumar2020}, wind power ramp events prediction \cite{DORADOMORENO2017428}, sea surface temperature prediction \cite{He2020}, study of the responses exhibited by plankton to fluid motions \cite{FUCHS2016109}, trends in solar radiation \cite{SILVA20101852} or simulation of extreme near shore sea conditions \cite{GOULDBY201415}. All these studies require a prior data collection and its adaptation to a specific format that allows the interpretation of them.

		Once quality and well-formed data are obtained, these can be used to extract information and build prediction models that explain the behavior of a certain problem. The choice of the appropriate model, either in engineering problems or in any other problem, is also an important factor in addition to data \cite{Alizadeh2019, Alizadeh2020}. \textcolor{blue}{In this sense, public and research organisms are increasingly promoting the use of open and robust datasets to boost policies coherent with the environmental exploitation, protection and conservation, as well as modelling tools available to the scientific community and decision-makers \cite{Manfren2020}}. Although Soft Computing (SC) and Machine Learning (ML) techniques have the ability to handle uncertainty in data and are extensively used for modelling purposes, the real challenge in modelling studies is due to the inadequacy of data, since the adequacy of the models depend mainly on the the quality of the information used, so that if a researcher does not have quality data there will be no quality models.

		Continuing with this line, special purpose software is usually developed to help researchers to advance in their studies related to energy and environmental modelling, becoming a great support for decision-making in the exploitation and protection of the environment. In \cite{Neeraj2020}, a software package in R called ``ForecastTB'' is developed for comparing the performance of distinct prediction methods, presenting the software as a stepping stone in ML automation modelling.
		In \cite{LO2015293}, an integrated simulation tool to optimise the design of bifacial solar panel with reflectors is presented. This tool can also be applied to study the efficiency of the solar cells. A framework for integrating information from offshore wind farms is implemented in \cite{NGUYEN2013150} in order to ease data interchange and enhance operation and maintenance practices. 
		In \cite{Roberta2020}, a new software tool named ``Storage LCA Tool`` is presented for comparing PCM - phase change materials- storage systems with conventional systems that do not involve energy storage, being beneficial for supporting decision-making on energy concepts for buildings.
		A risk assessment tool to improve safety standards and emergency management in onshore wind farms, is presented in \cite{ASTIASOGARCIA201648}. Raabe et al. \cite{RAABE2010213} developed two software tools, \textit{Model of Equilibrium of Bay Beaches} (MEPBAY) and \textit{Coastal Modelling System} (SMC), for supporting distinct operational levels of headland-bay beach in coastal engineering projects, and Motahhir et al. \cite{MOTAHHIR20199} developed an open hardware/software test bench for solar tracker.

		Marine energy prediction is currently a hot topic where meteorological data is used in. Marine Renewable Energy (MRE) is one of the most important renewable and sustainable energy sources available in our environment \cite{en12050787}, and it includes ocean thermal energy, marine tidal current energy and wave energy, among others. Its benefits and great potential \cite{ZEYRINGER20181281} make it one of the most relevant natural resources, playing a crucial role not only in the reduction of the emission of greenhouse gases but also in all other aspects involved in the difficult challenge of the transition to a low carbon footprint society \cite{en12091657, BREDE201344, Alizadeh2020a}. Wave energy exhibits a more stable power supply than wind energy and even solar energy. In recent years \textit{Wave Energy Converters} (WECs) \cite{FALCAO2010899} have been developed and widely installed, \textcolor{blue}{even arranged in array form \cite{Amini2020}}, to transform this wave energy into electricity, which can be injected into the electric network or supplied to existing offshore oil and gas platforms \cite{OLIVEIRAPINTO2019556} or seawater desalination plants \cite{FERNANDEZPRIETO2019546}, among others. WECs are mechanical devices that convert kinetic energy into electrical energy through either the vertical oscillation or the linear motion of waves. Nevertheless, waves are difficult to be characterised due to their stochastic nature, because of the influence of a large number of environmental factors that exert on them \cite{ochi1998}. As a consequence of this complexity, many aspects of WEC design, deployment and operation \cite{CROWLEY2018159, Abdelkhalik2016, 6898109} need a proper prediction of waves \cite{en11010011, Kaloop2020}, in order to maximise the wave energy extraction \cite{en80910370}. For this purpose WECs use wave \textit{flux of energy} ($F_e$) which can be calculated from the two most relevant wave parameters related to this aspect: \textit{significant wave height} ($H_s$) and \textit{wave energy period} ($T_e$).

		Currently, and as a support to traditional study procedures, SC and ML techniques \cite{Sang-Yong2014,Bishop:2006:PRM:1162264} are being widely used in numerous research fields related to classification, regression and optimisation tasks, obtaining significant improvements in the performance of the results, either in engineering \cite{Alizadeh2020}, energy or environmental problems \cite{Fi-John2019,Mosavi2019,GUO201816}. SC and ML methodologies can be used not only by experienced computer scientists but also by other researchers. For example, the well-known \textit{Waikato Environment for Knowledge Analysis} (WEKA) \cite{WEKA} software tool provides researchers with a wide collection of ML  algorithms. ML techniques have been already applied to tackle wave characterisation, accurately estimating $H_{s}$ and $T_{e}$ parameters \cite{DURANROSAL2017268, KUMAR2017605}, given that robustness of ML methods can tackle the previously explained difficulties in wave energy prediction. In \cite{Ali2020}, a reliable ML model based on multiple linear regression and covariant-weighted least square estimation for $H_s$ modelling is presented in order to predict significant wave height 30 minutes in advance. In \cite{Cornejo-Bueno2016}, an  approach for feature selection problems is developed and applied for $H_s$ and $F_e$ prediction in oceanic buoys, obtaining very good results. In \cite{Emmanouil2020}, a Bayesian Network system provides a helpful tool to support decision-making process of installation and maintenance operations in offshore wind farms using predictions of $H_s$, among others. In \cite{Shamshirband2020}, several ML methods are implemented and compared for the prediction of $H_s$ in the Persian Gulf, the extreme learning machine (ELM) providing the best results. The problem is that, in order to apply ML and SC techniques, it is essential to obtain datasets with relevant information about the issue under study, used to infer knowledge. Usually, these datasets are not publicly available in a friendly format, and their generation is the first step needed.

		The information to create these datasets related with MRE can be obtained from meteorological observations, but such information may be available in a inappropriate format and even contain missing values or measurements. Consequently, it is usually required to perform pre-processing tasks for improving the quality of the data, such as the replacement of missing values, outlier detection or data normalisation, among others. Furthermore, if more than one source of information is used to achieve a better characterisation of the problem under study \cite{JOHANSSON2015143, FERNANDEZ201544, Adams2010}, then a data integration process, denominated as the matching process in this document, has to be carried out by researchers to manually create the datasets with the needed information. Given that such process is of great relevance and has an extensive casuistry, the present work has been specially focused on it. Moreover, depending on the subject and the SC and ML technique to be applied, or even if the researcher considers other factors in order to enhance the performance obtained or have more in-depth conclusions, the datasets would have to be updated afterwards. In summary, many important details and different intermediate steps have to be considered when creating suitable datasets, specially when data integration is required, resulting in an extremely tedious task.

@@

			\item It makes the researcher focus on environmental modelling, without having to worry about the development of scripts or mechanical tasks, avoiding laborious pre-processing procedures, that imply a great deal of time and endeavour in early stages of the research.

@@

			\item It provides information about the quality and quantity of the data. SPAMDA allows preliminary studies of missing values (dates or measurements not recorded) in buoys managed by NDBC, so that the researcher can have an idea of the quality of the data recorded by the buoys and about their suitability for the intended purpose. In any case, SPAMDA allows data integration taking into account such missing values when needed by the user.

			\item Estimation of the amount of energy flux that can be produced at different prediction horizons: short-term, mid-term or long-term. Although this work does not focus on models performance, it should be taken into account that models tend to generalise worse with greater prediction horizons.

@@

			\item Possibility of selecting one or more reanalysis nodes near the localisation under study, which could provide a better description of the problem to achieve more accurate models.

			\item Although pre-processing is not the main objective of SPAMDA, the tool also provides some basic pre-processing filters on buoys measurements, such as normalisation and missing data recovery.

			\item It facilitates data management and well-organised storage of the datasets. Environmental studies in different geographical locations can be carried out by merely introducing and using other collected data.

@@

		Therefore, the functionalities and characteristics that SPAMDA offers make it a supporting tool for researchers, which could be used in applications related to coastal and ocean engineering, and also in marine energy prediction. In \cite{Kumar2020}, the estimation of energy supply sources in hybrid energy systems is based on the amount of energy that can be obtained by a marine energy system within a prediction horizon. Regulation of WECs to avoid malfunction or breakage, depending on the significant wave height and/or energy flux expected, as well as the possibility of reconfiguring them in order to maximise the wave energy extraction, is studied in \cite{CROWLEY2018159, Abdelkhalik2016}. The prediction of the energy that could be obtained from a certain maritime location is considered in \cite{OLIVEIRAPINTO2019556, FERNANDEZPRIETO2019546} in order to know whether it is convenient to install WECs as power supply in marine structures, such as offshore oil and gas platforms or seawater desalination plants. In \cite{Ali2019}, significant wave height forecasting is applied for decision-making in exploitation and environmental protection for the construction of marine energy storage plants, future strategies on renewable energy and coastal planning. Other examples of application are: design of offshore structures and ports \cite{CHATZIIOANNOU2017126}, decision-making and risk assessment about operational works in the sea \cite{DALGIC2015211}, security systems for structures or naval security \cite{Spaulding2020}.

@@

			\item NDBC belongs to the \textit{National Weather Service} (NWS) and operates and supports a network of marine and ocean buoys that record data. The mission of the network is to record marine and ocean meteorological data, such as $H_s$, dominant wave period, or wind speed and direction, among others.

			The buoys maintained by NDBC are located in coastal and offshore waters, and they are provided with specific sensors and devices which allow them to perform measurements. The information collected by the buoys is available on the NDBC website \cite{NOAA_1}, and it is divided into different groups. One of them corresponds to standard meteorological information of the historical data collected by each buoy, which can be downloaded as annual text files and whose format was adopted by NDBC since January 2007 \cite {NOAA_2}. These files contain hourly measurements per day from $00$:$50$ to $23$:$50$ UTC (Universal Time Coordinated) and from $23$:$50$ 31th December of the previous desired year to $22$:$50$ 31th December of the desired year. In Table \ref{tab:measurementsDescription}, a comprehensive measurement description and the corresponding units are provided as a summary for the reader. A fragment of one of these files, which contains the measurements collected during year $2017$ by the buoy identified as \textit{Station 46001} in NDBC, is shown in Fig. \ref{fig:fragmentAnnualTexFile}. Each column corresponds to a meteorological variable or attribute, and each row or instance corresponds to the values of the measurements collected by the buoy for each attribute at a specific date and time.

@@

				\caption{Measurements descriptions and units of each meteorological variable or attribute collected by the buoys (a detailed description can be found in NDBC website \cite{NOAA_4})}.
				\label{tab:measurementsDescription}
				\footnotesize
				\centering

				\begin{tabular}{ccm{7.0cm}@{\setlength{\tabcolsep}{0pt}}m{0.0cm}}

@@

					WDIR & degT & The direction the wind is coming from true North. \\

					\cellcolor{gray090}WSPD & \cellcolor{gray090} m/s & \cellcolor{gray090} The speed of the wind. \\

					GST &  m/s & Peak of gust speed. \\

					\cellcolor{gray090} WVHT & \cellcolor{gray090} m & \cellcolor{gray090} Significant wave height. \\

					DPD & sec & Dominant wave period (maximum wave energy). \\

					\cellcolor{gray090}APD & \cellcolor{gray090} sec & \cellcolor{gray090} Average wave period of all waves. \\

					MWD & degT & The direction from which the waves at the dominant period are coming. \\

					\cellcolor{gray090}PRES & \cellcolor{gray090} hPa & \cellcolor{gray090} Sea level pressure. \\

					ATMP & degC & Air temperature. &\\[0.10cm]

					\cellcolor{gray090}WTMP & \cellcolor{gray090} degC & \cellcolor{gray090} Sea surface temperature.\\

					DEWP & degC & Dewpoint temperature. \\

					\cellcolor{gray090}VIS & \cellcolor{gray090} nmi & \cellcolor{gray090} Visibility of the station. \\

					TIDE & ft & The water level. \\
